\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-04-15 honors_thesis/journal_paper/paper.tex
Ok, a couple of minor things.
Another two-line change
Made a bunch of revisions based on my read through and Andrew's feedback

In SEPSO future positions of the particles are speculated and evaluated in
parallel with the evaluation of current positions.  We show that a set of
speculative evaluation threads can be constructed such that the set will always
include the position and corresponding evaluation that would normally have
occurred at the next iteration of standard PSO.  Thus SEPSO performs two
iterations of PSO at once, allowing it to achieve better fitness faster.  Not
only can SEPSO can achieve better fitness in the same number of iterations, it
also reaches better fitness for the same number of total function evaluations
in large scale parallel environments.

@@

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\cite{mcnabb-2007-parallel-pso-using-mapreduce,
belal-2004-parallel-models-for-pso, chu-2006-intelligent-parallel-pso,
jin-2005-pso-antenna-designs, parsopoulos-2004-parallel-vector-evaluated-pso,
schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
\cite{koh-2006-parallel-asynchronous-pso,
mostaghim-2006-multi-objective-pso-on-grids,
venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
evaluation of the objective function can also be done in some cases, though
that is not an adaption of the PSO algorithm itself and thus is not the focus
of this paper.

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{evals_sphere}
  \caption{Function Sphere with a swarm with various swarm sizes.}
  \label{fig:evals-sphere}
\end{figure}

@@

%\begin{quote}
%SKIP SKIP
%
%No really, try reading without this! The idea of speculative execution can be applied to some types of optimization
%algorithms.  When the next sampling position of the algorithm does not depend
%on the value of the objective function at the current position,
%future positions can be speculated and evaluated in parallel with the current
%position.
%This is trivially true for optimization by random sampling, for example.
%The next random sample does not depend on the previous point nor on the value
%of the objective function at that point in any way.
%It depends only on a random number.
%There is no need to wait to see
%what objective function value was obtained at the previous point,
%evaluation of the object function at the next point can begin as soon as the next random number is drawn.
%In other optimization algorithms, the next position depends on the previous position, but not the
%value obtained there, a random walk for example.
%In this case
%the next position to be evaluated
%depends only on the previous position and a random number.
%In this case, as it was in the case of random sampling,
%the object function evaluation of the next position can be done in parallel
%because it 
%does not depend on the \emph{value} of the objective function at the previous position.
%There are an infinite number of possible next positions, but only one given the draw of the random number.
%
%In still other cases, the next position to be evaluated may depend on the current position and some
%discrete valued (binary, for example) test on the value obtained at the current position.
%For example, consider a greedy random walk, in which the state only advances if a better
%position is found.
%That is, if the current value is better than the previous value,
%use the sum of the current position and the random number as the next position.
%Otherwise, computer the next position using the sum of the previous position and the random number.
%In this case the objective function can be evaluated at three positions:
%(1) the value of the objective function at the current position,
%(2) the value at the position if the test turns out to be true and (3) the value at the position
%if the test turns out to be false; all in parallel.
%In this example it is important to note that we do not need the value of the objective function at the current
%point to to reduce the infinite number of possible next positions to just two alternatives.
%Choosing between these two alternatives can be done post-hoc, after the objective function value
%at the current position and at the two alternative are computed.
%For all but the most computationally-trivial objective functions, this approach saves almost half of the
%``wall-clock'' time.
%This approach has been applied to simulated annealing~\cite{witte-1991-parallel-simulated-annealing-speculative} which like the greedy random walk,
%depends on positions and a binary test.
%
%This approach can not be used in gradient methods where the
%step size for sampling typically depends on the gradient and therefor the position and 
%a set of infinitely many possible \emph{values} outcomes
%obtained at the current position (not binary or even finite).
%\end{quote}

% Use this later?
% Skip this??
% 
% Speculative evaluation is possible in PSO because the motion of particles
% depends only on the \emph{positions} of the particle and its best neighbor, not
% the \emph{value} of the objective function at those positions.  We do not have
% to wait for a particle's function evaluation to complete in order to know where
% it might move to next, so we can compute the set of all possible next positions
% and send them to be evaluated along with the particle's current position.  Once
% the current position of each particle has been evaluated, we can determine
% which of the speculatively evaluated positions matches the branch taken by the
% particle.  We have thus evaluated two iterations at once.  This can easily get
% unwieldy if the set of next possible positions for each particle is large, but
% a wise choice of topology limits that set to a reasonable size.
% 
% But the following does need some intro

In this paper we show that the results of standard PSO can be reproduced
\emph{exactly}, two iterations at a time, using a speculative approach adapted
from speculative execution. We prove that the standard PSO equation can be
factored such that a set of speculative positions can be found which will
\emph{always} include the position computed in the next iteration.  By
computing the value of the objective function for each of the speculative
positions at the same time the algorithm evaluates the objective function for
the current position, it is possible to know the objective function values for
both the current and the next iteration at the same time.  We demonstrate this
principle by implementation and show that it produces exactly the same results
as standard PSO, but two iterations at a time.  The resulting implementation
runs efficiently on large clusters where the number of processors is much
larger than a typical or reasonable number of particles, producing better
results in less ``wall-clock'' time and, surprisingly, with fewer total
function evaluations.

@@

In this form the important point to notice is that there are only 7 values
(for this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
\caseset\}$ and that none of them depend upon $f(\ppos_t)$ or any other
objective function evaluation at iteration $t$ (note also that while there are
random numbers in the equation, they are assumed fixed once drawn for any
particular particle at a specific iteration).  Thus PSO has been refactored
such that the algorithm can begin computing all 7 of the objective function
evaluations potentially needed in iteration $t+1$ \emph{before} $f(\ppos_t)$
is computed.  Once the evaluation of $f(\ppos_{t})$ is completed for all
particles only one of the indicator functions $\indic_{t+1}^{\casegen}$ will
be set to 1.  Although this speculative algorithm computes
$\ofunc(\specpos_{t+1}^{\casegen})$ for all $\casegen \in \caseset$, even
those for which $\indic_{t+1}^{\casegen} = 0$, these extra computations will
be ignored, and might just as well never have been computed.  We call the set
of computations $\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$
``speculative children'' because only one of them will be needed.

@@

In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
master, keeps track of all necessary information, and all other machines are
merely used to evaluate the objective function at various positions as directed
by the master~\cite{belal-2004-parallel-models-for-pso}.  To perform
speculative evaluation in such an architecture, the master generates the
positions to evaluate speculatively as in \eqref{eq:pos2update}.  After having
the slaves evaluate the objective function at all necessary positions, the
master then decides which position to accept for each particle, as in
\eqref{eq:val2updatelong}.  The outline of the procedure is given in
\alg{centralized}.

\begin{algorithm}
  \caption{Speculative Evaluation in a Centralized PSO}
  \label{alg:centralized}
  \begin{algorithmic}[1]
	\STATE Move all $\p_{t-1}$ to $\noeval{\p}_t$ using
	  \eqref{eq:velupdate}~and~\eqref{eq:posupdate}
	\STATE For each $\noeval{\p}_t$, get its neighbors $\noeval{\nset}_t$ and
	  generate $\noeval{\sset}_{t+1}$ according to
	  \eqref{eq:pos2update}.
	\STATE Evaluate all $\noeval{\p}_t$ and $\noeval{\sset}_{t+1}$ in parallel
	\STATE Update personal best for each $\noeval{\p}_t$ and
	  $\noeval{\s}_{t+1}$, creating $\nonbest{\p}_t$ and $\nonbest{\s}_{t+1}$
	\STATE Update neighborhood best for each $\nonbest{\p}_t$, creating
	  $\pset_t$
	\FORALL{$\p_t$}
	\STATE Pick $\nonbest{\s}_{t+1}$ from $\nonbest{\sset}_{t+1}$ that matches
	  the branch taken by $\p_t$ according to
	  \eqref{eq:val2updatelong}.
	\STATE Pass along personal and neighborhood best values obtained by $\p_t$,
	  making $\nonbest{\p}_{t+1}$
	\ENDFOR
	\STATE Update neighborhood best for each $\nonbest{\p}_{t+1}$, creating
	  $\pset_{t+1}$
	\STATE Repeat from Step 1 until finished
  \end{algorithmic}
\end{algorithm}

Given a set of particles at iteration $t-1$ (perhaps which have just been
initialized), the master must move each particle using
\eqref{eq:velupdate} and \eqref{eq:posupdate} to obtain the set
$\noeval{\pset}_t$.  For each particle $\noeval{\p}_t$, the master must get its
set of neighbors $\noeval{\nset}_t$ and use their positions, along with the
position of $\noeval{\p}_t$, to enumerate all possible combinations of $\pbest$
and $\nbest$ that particle $\noeval{\p}_t$ could end up with (as shown in
Table~\ref{tab:evals}).  Each of those combinations defines a speculative
particle, $\s_t$, that can be moved using \eqref{eq:pos2update} to
obtain $\noeval{\s}_{t+1}$.  The result is a set of particles
$\noeval{\pset}_t$, and for each particle a set of speculative children
$\noeval{\sset}_{t+1}$, which can all be evaluated at once.

The master then has the slaves evaluate the particles.  Once all particles,
speculative and original, have been evaluated and the values reported to the
master, the master determines which speculative child of each particle was the
correct one.

This is done by first updating each particle's personal and neighborhood bests.
Mathematically, this corresponds to the evaluation of an indicator function
similar to that found in \eqref{eq:deficasexn}.  In practice, this is
done by moving each $\noeval{\p}_t$ to $\nonbest{\p}_t$ and each speculative
particle $\noeval{\s}_{t+1}$ to $\nonbest{\s}_{t+1}$ by associating the value
determined by the slave to the particle it came from, and updating that
particle's personal best if the value reported is better than its previous
personal best value.  Then each $\nonbest{\p}_t$ must be updated to $\p_t$ by
updating its neighborhood best position and value with information about each
of its neighbors.  This is simply the original PSO algorithm, and corresponds
to steps 1--5 in \alg{centralized}.

Then, with the set $\pset_t$, the branch that was actually taken by each
particle can be determined, and the correct speculative child can be chosen.
The branches depend entirely on whether or not the personal best of $p_t$ was
updated and which particle, if any, updated the particle's neighborhood best.
Both of those items are now known, and the child that matches the branch is
selected as per \eqref{eq:val2updatelong} (step 7 in
\alg{centralized}).

@@

In a distributed parallel PSO algorithm, individual processors not only perform
evaluations of particles, but also their movement.  The information for each
particle is not held by a central machine that directs the algorithm; instead,
each processor has the information for the particle or particles that it is in
charge of and must perform the steps of the algorithm for those particles.
Messages such as values and positions for the neighborhood best are sent
between processors.  There may still be some machine that collects information
from all of the particles and outputs the result of the algorithm, though that
machine's importance is much less than in centralized algorithms.

@@

The speculative algorithm actually performs two iterations of PSO at each
``iteration,'' so we instead call each ``iteration'' a round of evaluations.

To use the same number of processors (and hence evaluations) for each
algorithm, speculative evaluation must use a smaller swarm size than standard
PSO.  For each of the topologies we used, a particle has three neighbors
including itself.  As shown in Table~\ref{tab:evals}, this results in $7$
speculative evaluations per particle.  With one evaluation needed for the
original, non-speculative particle, we have $8p_s$ evaluations for every two
iterations, where $p_s$ is the number of particles in the speculative swarm.
The extra evaluations done in speculative evaluation would instead be particles
in regular PSO, so we compare swarms of size $p_s$ in speculative evaluation
with swarms of size $8p_s$ in regular PSO.

Where the Complete topology would normally be used, we use a Random topology in
our speculative algorithm, as Complete leads to an explosion in the number of
speculative evaluations.  If speculative evaluation were not being performed,
it is possible that the Complete topology would be used.  However, the Complete
topology also requires a very large amount of interprocessor communication in
parallel PSO, so it is still quite possible that Random would be used even with
standard PSO.  But, to be fair in our comparisons, we compare to the standard
PSO algorithm using both the Random topology with two neighbors and the
Complete topology.  All functions have 20 dimensions, and all results shown are
averages over 20 runs.  We use 240 processors in our experiments (except where
otherwise noted), so our speculative algorithm uses 30 particles and standard
PSO uses 240 particles.  For ease of labeling, we call our speculative
algorithm SEPSO and standard PSO just PSO, with the topology following.

@@

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optimum.  Griewank has a global optimum with a value of 0, and
sometimes the swarm finds the optimum and sometimes it doesn't.  Instead of
showing average function value at each iteration, a more accurate plot for
Griewank shows the percent of runs that have found the global optimum by each
iteration.  We show results in \fig{basic-griewank2} for swarms of size 30 and
240 using the Ring topology.  One can see in the figure that the SEPSO
algorithm finds the optimum faster than the original PSO.  However, because the
swarm size is so small, SEPSO gets stuck a little less than half of the time.

@@

Because our speculative algorithm got stuck with such a small swarm size, we
ran a set of experiments on Griewank with 800 processors, giving swarms of size
100 and 800 instead of 30 and 240.  \fig{basic-griewank3} shows the results.
\fig{basic-griewank3} is very similar to \fig{basic-griewank2}, except that
speculative evaluation finds the global optimum essentially 100\% of the time.
Thus with a few more processors, our algorithm finds the optimum twice as fast
as the original PSO without a significant loss of accuracy.

@@

To use the extra speculative evaluations, instead of choosing the speculative
child that matches the branch that the original PSO would have taken (the
``behavior preserving'' branch), we take the child that has the best value.
The methodology is exactly the same as above except for the process of choosing
which speculative child to accept.  The only change needed in \alg{centralized}
is in step 7, where the $\noeval{\s}_{t+1}$ with the best value is chosen from
$\noeval{\sset}_{t+1}$ instead of with the matching branch.

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank4}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing Pick Best with previous results.}
  \label{fig:griewank-pickbest}
\end{figure}

@@

\begin{table}[ht]
  \caption{Five Branches to Consider for Statistics}
  \label{tab:branches}
  \centering
  \begin{tabular}{lcc}
	Identifier&Source of $\pbest$ update&Source of $\nbest$ update\\
	\hline
	\hline
	$\casepn$&No update&No update\\
	\hline
	$\casexn$&Self&No update\\
	\hline
	$\casexx$&Self&Self\\
	\hline
	$\casepl$&No update&Some Neighbor\\
	\hline
	$\casexl$&Self&Some Neighbor\\
	\hline
  \end{tabular}
\end{table}

@@

Row 1 in Table~\ref{tab:branches}, \casepn, corresponds to the phenomenon
commonly called stagnation in the literature, but only with a single particle.
As an interesting aside, we found that complete stagnation (where the entire
swarm is stagnant) practically never occurs in PSO, though frequently large
percentages of the swarm are stagnant.  Iterations where the entire swarm is
stagnant average less than 2 in 1000 iterations, unless the swarm converges
beyond machine precision.  When graphs of the best function value flat-line,
what really has happened is a contraction of the particles' velocity, not
stagnation in the technical sense.

@@

From Table~\ref{tab:stats} we can see that with the Random topology, we can
pick the first branch, corresponding to stagnation, and be right around 70\% of
the time.  With the Ring topology, we would be right 50\% of the time.
Branches \casepN and \casexN really correspond to several actual branches, as
all of the neighbors are lumped together.  The 20\% probability of taking
branch \casepN with the Ring topology can be split into two branches, as there
are only two neighbors.  It also turns out that the neighbor that last updated
the neighborhood best is the most likely to update it next time, so keeping
track of that information could be fruitful in trying to predict that branch.

The statistics for the Complete topology are less promising, as there are 240
neighbors that branch \casepN splits into, instead of two.  Pruning does,
however, allow for the possibility of using the Complete topology in
speculative evaluation while avoiding the explosion in the number of
evaluations it would otherwise entail.  Though pruning makes its use possible,
we still found the communication costs in our parallel environment prohibitive,
so we do not present any results for the Complete topology here.

@@

\subsection{What if we guessed wrong?}
\label{sec:wrong}

When not all of the branches are evaluated, there is some probability of not
evaluating the branch that was actually taken by the original particle.  We 
need to decide what to do in this case.

One possibility is to leave the particle as it is, not accepting any of the
speculative evaluations, because none of them were correct.  This leads to
particles being at different iterations, as some particles guess correctly
while others do not.  Thus we lose exact compatibility with the original PSO,
though this particular relaxation is nothing new; asynchronous adaptations of
PSO do the same thing.  As an aside, it is equivalent in this case to simply
increment the iteration number of particles which fail to correctly predict
their branch.  This keeps the iteration number constant across all particles,
simplifying the work needed to be done in determining neighbors when dynamic
topologies are involved.  We call this technique Social Promotion.

@@

There are a large number of ways to prune speculative evaluations.  We
experimented with several, but present results only for one possible pruning.
The pruning we present uses only the two most probable branches in the Random
topology.  As seen in Table~\ref{tab:stats}, these are the two branches where
the $\nbest$ value was not updated---\casepn and \casexn.  Those branches are
also convenient in that no messages are needed from neighbors in order to
produce the positions of the speculative particles---in distributed frameworks
using several rounds of communication (see \sec{distributed}), one of the
rounds of communication can be dropped entirely.

@@

In \fig{sphere-pruned}, we show the results of this simple pruning on the
function Sphere.  As can be seen, the larger number of particles that pruning
allows results in better results when picking the best speculative child.  
When rejecting all speculative children no real improvement is seen over the
original PSO algorithm.

@@

The question of which branches to take in this infinite tree is an intriguing
one that we can only begin to explore here.  If the branch corresponding to
stagnation has a 75\% chance of being taken, as in the Random topology with
most functions, we could speculate three iterations ahead on that branch and
still have a 42\% chance of predicting correctly.  However, intuition would say
that perhaps it is better to hope that the particle is productive instead of
stagnant, so a branch where the particle updates its personal best might be
more fruitful to try.  In our experiments we try just one of the countless
possibilities, but that one turns out to work very well.  More work is needed
to compare the different branching possibilities on various functions.

The combination of branches we tried uses seven speculative evaluations per
particle, matching the swarm size of the original speculative algorithm.  The
seven evaluations we used corresponded to several iterations of branches
\casepn and \casexn.  Two of the evaluations were just one iteration ahead,
four were two iterations ahead, and one was three iterations ahead.  The
evaluations that were one iteration ahead were branches \casepn and \casexn;
those that went two iterations ahead were formed by taking either branch
\casepn or \casexn and then branch \casepn or \casexn; and the evaluation going
three iterations ahead followed branch \casepn on all three iterations.

@@

Rastrigin, on the other hand, is a function for which increasing the swarm size
up to 4000 particles still shows improvements in the performance of the
algorithm.  With such functions, the smaller swarm size required by speculative
evaluation is not able to produce enough exploration to perform better than
standard PSO at the swarm sizes we were able to experiment with.

@@

We have speculated that our Pick Best method simply draws a number of samples
from the next iteration of the sampling distribution of PSO and keeps one of
them.  It would then be expected that the sampling distribution of that method
is very similar to the distribution of the original PSO.  This remains to be
proven, or even demonstrated empirically.

>>>>

In SEPSO future positions of the particles are speculated and evaluated in
parallel with the evaluation of current positions.  We show that a set of
speculative evaluations can be constructed such that the set will always
include the position and corresponding evaluation that would normally have
occurred at the next iteration of standard PSO.  Thus SEPSO performs two
iterations of PSO at once, allowing it to achieve better fitness faster.

@@

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\cite{chu-2006-intelligent-parallel-pso, jin-2005-pso-antenna-designs,
parsopoulos-2004-parallel-vector-evaluated-pso,
schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
\cite{mostaghim-2006-multi-objective-pso-on-grids,
venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
evaluation of the objective function can also be done in some cases, though
that is not an adaption of the PSO algorithm itself and thus is not the focus
of this paper.

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{evals_sphere}
  \caption{Function Sphere with various swarm sizes. Error bars show median and
  10th and 90th percentiles.}
  \label{fig:evals-sphere}
\end{figure}

@@

In this paper we show that the results of standard PSO can be reproduced
\emph{exactly}, two iterations at a time, using a speculative approach adapted
from speculative execution. We prove that the standard PSO equation can be
factored such that a set of speculative positions can be found which will
\emph{always} include the position computed in the next iteration.  By
computing the value of the objective function for each of the speculative
positions at the same time the algorithm evaluates the objective function for
the current position, it is possible to know the objective function values for
both the current and the next iteration at the same time.  We demonstrate this
principle by implementation and show that it produces exactly the same results
as standard PSO, but two iterations at a time.  The resulting implementation
runs efficiently on large clusters where the number of processors is much
larger than a typical or reasonable number of particles, producing better
results in less ``wall-clock'' time.

@@

In this form the important point to notice is that there are only 7 values (for
this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
\caseset\}$ and that none of them depend upon $f(\ppos_t)$ or any other
objective function evaluation at iteration $t$ (note also that while there are
random numbers in the equation, they are assumed fixed once drawn for any
particular particle at a specific iteration).  Thus PSO has been refactored
such that the algorithm can begin computing all 7 of the objective function
evaluations potentially needed in iteration $t+1$ \emph{before} $f(\ppos_t)$ is
computed.  Once the evaluation of $f(\ppos_{t})$ is completed for all particles
only one of the indicator functions $\indic_{t+1}^{\casegen}$ will be set to 1;
hence only one of the positions $\specpos_{t+1}^\casegen$ will be kept.

Although this speculative algorithm computes
$\ofunc(\specpos_{t+1}^{\casegen})$ for all $\casegen \in \caseset$, even those
for which $\indic_{t+1}^{\casegen} = 0$, these extra computations will be
ignored, and might just as well never have been computed.  We call the set of
computations $\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$
``speculative children'' because only one of them will be needed.

@@

In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
master, keeps track of all necessary information, and all other machines are
merely used to evaluate the objective function at various positions as directed
by the master~\cite{belal-2004-parallel-models-for-pso}.  To perform
speculative evaluation in such an architecture, the master generates the
positions to evaluate speculatively as in \eqref{eq:defpcasegen}.  After having
the slaves evaluate the objective function at all necessary positions, the
master then decides which position to accept for each particle, as in
\eqref{eq:val2updatelong}.  The outline of the procedure is given in
\alg{centralized}.

\begin{algorithm}
  \caption{Speculative Evaluation in a Centralized PSO}
  \label{alg:centralized}
  \begin{algorithmic}[1]
	\STATE Move all $\p_{t-1}$ to $\noeval{\p}_t$ using
	  \eqref{eq:velupdate}~and~\eqref{eq:posupdate}
	\STATE For each $\noeval{\p}_t$, get its neighbors $\noeval{\nset}_t$ and
	  generate $\noeval{\sset}_{t+1}$ according to
	  \eqref{eq:defpcasegen}.
	\STATE Evaluate all $\noeval{\p}_t$ and $\noeval{\sset}_{t+1}$ in parallel
	\STATE Update personal best for each $\noeval{\p}_t$ and
	  $\noeval{\s}_{t+1}$, creating $\nonbest{\p}_t$ and $\nonbest{\s}_{t+1}$
	\STATE Update neighborhood best for each $\nonbest{\p}_t$, creating
	  $\pset_t$
	\FORALL{$\p_t$}
	\STATE Pick $\nonbest{\s}_{t+1}$ from $\nonbest{\sset}_{t+1}$ that matches
	  the branch taken by $\p_t$ according to
	  \eqref{eq:val2updatelong}.
	\STATE Pass along personal and neighborhood best values obtained by $\p_t$,
	  making $\nonbest{\p}_{t+1}$
	\ENDFOR
	\STATE Update neighborhood best for each $\nonbest{\p}_{t+1}$, creating
	  $\pset_{t+1}$
	\STATE Repeat from Step 1 until finished
  \end{algorithmic}
\end{algorithm}

Given a set of particles at iteration $t-1$ (perhaps which have just been
initialized), the master must move each particle using \eqref{eq:velupdate} and
\eqref{eq:posupdate} to obtain the set $\noeval{\pset}_t$.  For each particle
$\noeval{\p}_t$, the master must then get its set of neighbors
$\noeval{\nset}_t$ and use their positions, along with the position of
$\noeval{\p}_t$, to calculate all possible values of
$\specpos_{t+1}^{\casegen}$, using \eqref{eq:defpcasegen}.  These positions,
along with the original particle's associated information (such as values for
$\pbest$ and $\nbest$), define a set of speculative children,
$\noeval{\sset}_{t+1}$.  The master than has a set of particles
$\noeval{\pset}_t$, and for each particle a set of speculative children
$\noeval{\sset}_{t+1}$, which can all be evaluated at once.

The master then has the slaves evaluate the particles.  Once all particles,
speculative and original, have been evaluated and the values reported to the
master, the master determines which speculative child of each particle was the
correct one.  Mathematically, this corresponds to the evaluation of an
indicator function similar to that found in \eqref{eq:deficasexn}.  In
practice, this is done first by updating each particle's $\pbest$, if
necessary, then by updating the particle's $\nbest$ with information from the
particle's neighbors.  This is simply the original PSO algorithm, and
corresponds to steps 1--5 in \alg{centralized}.  Given the updates to $\pbest$
and $\nbest$, the case from Table~\ref{tab:evals} can be determined, as per
\eqref{eq:val2updatelong}.  The child with the matching case is kept, and all
other speculative children are discarded (step 7 in \alg{centralized}.

@@

In a distributed parallel PSO algorithm, individual processors not only perform
evaluations of particles, but also their movement.  The information for each
particle is not held by a central machine that directs the algorithm; instead,
each processor has the information for the particle or particles that it is in
charge of and must perform the steps of the algorithm for those
particles~\cite{mcnabb-2007-parallel-pso-using-mapreduce}.  Messages such as
values and positions for the neighborhood best are sent between processors.
There may still be some machine that collects information from all of the
particles and outputs the result of the algorithm, though that machine's
importance is much less than in centralized algorithms.

@@

The speculative algorithm actually performs two iterations of PSO at each
``iteration,'' so we instead call each ``iteration'' a ``round of
evaluations''.

To use the same number of processors (and hence evaluations) for each
algorithm, speculative evaluation must use a smaller swarm size than standard
PSO.  For each of the topologies we used, a particle has three neighbors
including itself.  As shown in Table~\ref{tab:evals}, this results in $7$
speculative evaluations per particle.  With one evaluation needed for the
original, non-speculative particle, we have $8p$ evaluations for every two
iterations, where $p$ is the number of particles in the speculative swarm.
The extra evaluations done in speculative evaluation would instead be particles
in regular PSO, so we compare swarms of size $p$ in speculative evaluation
with swarms of size $8p$ in regular PSO.

Where the Complete topology would normally be used, we use a Random topology in
our speculative algorithm, as Complete leads to an explosion in the number of
speculative evaluations.  If speculative evaluation were not being performed,
it is possible that the Complete topology would be used.  However, the Complete
topology also requires a very large amount of interprocessor communication in
distributed PSO, so it is still quite possible that Random would be used even
with standard PSO.  But, to be fair in our comparisons, we compare to the
standard PSO algorithm using both the Random topology with two neighbors and
the Complete topology.  All functions have 20 dimensions, and all results shown
are averages over 20 runs.  We use 240 processors in our experiments (except
where otherwise noted), so our speculative algorithm uses 30 particles and
standard PSO uses 240 particles.  For ease of labeling, we call our speculative
algorithm SEPSO and standard PSO just PSO, with the topology following.

@@

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optimum~\cite{bratton-2007-defining-a-standard-for-pso}.  Griewank
has a global optimum with a value of 0, and sometimes the swarm finds the
optimum and sometimes it doesn't.  Instead of showing average function value at
each iteration, a more informative plot for Griewank shows the percent of runs
that have found the global optimum by each iteration.  We show results in
\fig{basic-griewank2} for swarms of size 30 and 240 using the Ring topology.
One can see in the figure that the SEPSO algorithm finds the optimum faster
than the original PSO.  However, because the swarm size is so small, SEPSO gets
stuck a little less than half of the time.

@@

Because the speculative algorithm got stuck with such a small swarm size, we
ran a set of experiments on Griewank with 800 processors, giving swarms of size
100 and 800 instead of 30 and 240.  \fig{basic-griewank3} shows the results.
\fig{basic-griewank3} is very similar to \fig{basic-griewank2}, except that
speculative evaluation finds the global optimum essentially 100\% of the time.
Thus with a few more processors, our algorithm finds the optimum twice as fast
as the original PSO without a significant loss of accuracy.

@@

To use the extra speculative evaluations, instead of choosing the speculative
child that matches the branch that the original PSO would have taken, we take
the child that has the best value.  The methodology is exactly the same as
above except for the process of choosing which speculative child to accept.
The only change needed in \alg{centralized} is in step 7, where
$\noeval{\s}_{t+1}$ with the best value is chosen from $\noeval{\sset}_{t+1}$
instead of with the matching branch.

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank4}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing Pick Best with previous results.}
  \label{fig:griewank-pickbest}
\end{figure}

@@

\begin{table}[ht]
  \caption{Five Branches to Consider for Statistics}
  \label{tab:branches}
  \centering
  \begin{tabular}{lcc}
	Identifier&Source of $\pbest$ update&Source of $\nbest$ update\\
	\hline
	\hline
	$\casepn$&No update&No update\\
	\hline
	$\casexn$&Self&No update\\
	\hline
	$\casexx$&Self&Self\\
	\hline
	$\casepN$&No update&Some Neighbor\\
	\hline
	$\casexN$&Self&Some Neighbor\\
	\hline
  \end{tabular}
\end{table}

@@

Row 1 in Table~\ref{tab:branches}, \casepn, corresponds to the phenomenon
commonly called stagnation in the literature, but with only a single particle.
As an interesting aside, we found that complete stagnation (where the entire
swarm is stagnant) practically never occurs in PSO, though frequently large
percentages of the swarm are stagnant.  Iterations where the entire swarm is
stagnant average less than 2 in 1000 iterations, unless the swarm converges
beyond machine precision.  When graphs of the best function value flat-line,
what really has happened is a contraction of the particles' velocity, not
stagnation in the technical sense.

@@

From Table~\ref{tab:stats} we can see that with the Random topology, we can
pick the first branch, corresponding to stagnation, and be right around 70\% of
the time.  With the Ring topology, we would be right 50\% of the time.
Branches \casepN\ and \casexN\ really correspond to several actual branches, as
all of the neighbors are lumped together.  The 20\% probability of taking
branch \casepN\ with the Ring topology can be split into two branches, as there
are only two neighbors.  It also turns out that the neighbor that last updated
the neighborhood best is the most likely to update it next time, so keeping
track of that information could be fruitful in trying to predict that branch.

The statistics for the Complete topology are less promising, as there are 240
neighbors that branch \casepN\ splits into, instead of two.  Pruning does,
however, allow for the possibility of using the Complete topology in
speculative evaluation while avoiding the explosion in the number of
evaluations it would otherwise entail.  Though pruning makes its use possible,
we still found the communication costs in our parallel environment prohibitive,
so we do not present any results for the Complete topology here.

@@

\subsection{Recovering from Pruning Too Much}
\label{sec:wrong}

When not all of the branches are evaluated, there is some probability of not
evaluating the branch that was actually taken by the original particle.  There
are a few possibilities for recovery in this case.  One is to leave the
particle as it is, not accepting any of the speculative evaluations, because
none of them were correct.  This leads to particles being at different
iterations, as some particles guess correctly while others do not.  Thus we
lose exact compatibility with the original PSO, though this particular
relaxation is nothing new; asynchronous adaptations of PSO do the same
thing~\cite{koh-2006-parallel-asynchronous-pso}.  As an aside, it is equivalent
in this case to simply increment the iteration number of particles which fail
to correctly predict their branch.  This keeps the iteration number constant
across all particles, simplifying the work needed to be done in determining
neighbors when dynamic topologies are involved.  We call this technique Social
Promotion.

@@

There are a large number of ways to prune speculative evaluations.  We
experimented with several, but present results only for one possible pruning.
The pruning we present uses only the two most probable branches in the Random
topology.  As seen in Table~\ref{tab:stats}, these are the two branches where
the $\nbest$ value was not updated: \casepn\ and \casexn.  Those branches are
also convenient in that no messages are needed from neighbors in order to
produce the positions of the speculative particles---in distributed frameworks
using several rounds of communication (see \sec{distributed}), one of the
rounds of communication can be dropped entirely.

@@

In \fig{sphere-pruned}, we show the results of this simple pruning on the
function Sphere.  As can be seen, the larger number of particles that pruning
allows leads to better results when picking the best speculative child.  When
using Social Promotion no real improvement is seen over the original PSO
algorithm.

@@

The question of which branches to take in this infinite tree is an intriguing
one that we can only begin to explore here.  If the branch corresponding to
stagnation has a 75\% chance of being taken, as in the Random topology with
most functions, we could speculate three iterations ahead on that branch and
still have a 42\% chance of predicting correctly.  However, intuition would say
that perhaps it is better to hope that the particle is productive instead of
stagnant, so a branch where the particle updates its personal best might be
more fruitful to try.  In our experiments we try just one of the countless
possibilities, but one that turns out to work very well.  More work is needed
to compare the different branching possibilities on various functions.

The combination of branches we tried uses seven speculative evaluations per
particle, matching the swarm size of the original speculative algorithm.  The
seven evaluations we used corresponded to several iterations of branches
\casepn\ and \casexn.  Two of the evaluations were just one iteration ahead,
four were two iterations ahead, and one was three iterations ahead.  The
evaluations that were one iteration ahead were branches \casepn\ and \casexn;
those that went two iterations ahead were formed by taking either branch
\casepn\ or \casexn and then branch \casepn\ or \casexn; and the evaluation
going three iterations ahead followed branch \casepn\ on all three iterations.

@@

Rastrigin, on the other hand, is a function for which increasing the swarm size
up to 4000 particles still shows improvements in the performance of the
algorithm.  With such functions, the smaller swarm size required by speculative
evaluation does not produce enough exploration to perform better than standard
PSO at the swarm sizes we were able to experiment with.

@@

We have speculated that our Pick Best method simply draws a number of samples
from the next iteration of the sampling distribution of PSO and keeps one of
them.  It would then be expected that the sampling distribution of that method
is very similar to the distribution of the original
PSO~\cite{poli-2008-sampling-distribution-of-pso}.  This remains to be proven,
or even demonstrated empirically.
